From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)

### 2025-04-13 analysis/paper/paper.qmd
References updated on main doc

The proportions of local vs non-local raw materials discussed thus far have focused greatly on group mobility, which is linked with residential mobility and occupation duration. Residential mobility (e.g., foraging range size and frequency of residential moves) is a defining characteristic of hunter-gatherer societies and plays a central role in their lifeways  (Kelly, 1995). It is a key factor driving changes in lithic assemblages (Brown, 1999; Oestmo, 2017; Roth & Dibble, 1998) as it significantly impacts the cost and use of artefacts and raw materials since they affect the relation between the distance of raw material sources to sites (Kuhn, 2004). As such, different raw material and technological patterns have been observed between short-term and long-term occupations.

Short-term occupations typically rely on transported (non-local) toolkits (Torrence, 1983) and raw materials (Surovell, 2009), show higher reduction intensity, and an increased tools-to-debitage ratio in the assemblages (Grove et al., 2023). Long-term occupations typically rely on provisioning sites with locally available and suitable raw materials, while non-local materials are gradually used and discarded (Grove et al., 2023; Surovell, 2009). Non-local raw material cores are highly reduced and extensively exploited (Kuhn, 2004; Roth & Dibble, 1998) and exhibit higher blank-to-core ratios (Roth & Dibble, 1998).

Altogether, different studies highlight significant variability in lithic assemblages and raw material proportions, reflecting different facets of hunter-gatherer technological organisation. This variability appears between behaviourally and ethnically similar sites (Bamforth, 1986), across distinct chrono-cultural periods, and even within a single site among various raw materials. To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production— procurement, distribution, use, maintenance, and discard—through interconnected approaches that combine lithic technology and raw material analyses.

@@

Vale Boi is located on the western coast of the Algarve region, southern Portugal, within a small valley extending \~2 km southward to the Atlantic Ocean (\@fig-map a). The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters (\@fig-map b). Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter (\@fig-map c), which were excavated between 2000 and 2019 (Bicho et al., 2007, 2012, 2013; Cascalheira et al., 2017; Cascalheira & Bicho, 2013; Manne et al., 2012). Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data (Belmiro et al., 2025), the Terrace and Shelter areas were selected for the current study.

@@

The Terrace area preserves occupations ranging from the UP to the Early Neolithic. While excavations started in 2000, an additional 8-square-meter trench was opened in 2012 to refine the stratigraphic sequence and assess the presence of earlier UP deposits. Between 2012 and 2016, eight primary litho-stratigraphic units were identified. The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) dated between approximately c. 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c. 26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP (Belmiro et al., 2021; Cascalheira et al., 2017; Cascalheira & Bicho, 2013).

The Shelter area contains four main lithostratigraphic units, with occupations from the Magdalenian, Solutrean, and Gravettian periods. The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers A to C), dated to approximately c. 24-22 ka cal BP. These occupation levels were found beneath blocks of limestone that likely collapsed from the rock shelter ceiling, probably shortly after the LGM (Cascalheira, 2010; Cascalheira et al., 2017).

The site’s lithic assemblages represent all stages of stone tool production, with raw material use varying by occupation. Technological analyses of these assemblages have provided insights into the adaptive strategies of UP hunter-gatherers at Vale Boi, revealing technological shifts across different cultural phases. Given the chosen sample for this study (see **Section 2.2. Datasets and analysis**), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter (Cascalheira, 2010; Cascalheira et al., 2013) and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace (Belmiro, 2020; Belmiro et al., 2021). The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.

The lithic analysis of the Solutrean levels from the Shelter (levels A-C) shows a high number of lithic artifacts, with over 22,000 artifacts analysed (including chips which correspond to \~60% of the assemblage and shatter \~17-19%). The volumes introduced into the site were mostly cortical, given the high frequencies of cortex in blanks and retouched tools. The knapping strategies are dominated by unidirectional and bidirectional prismatic cores, with frequent core maintenance products. Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blanks are the most common retouched tools. Stemmed and winged points, laurel leaves, and shouldered points were also identified (Cascalheira et al., 2013).

The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to \~65-70% of the assemblage and shatter \~20%), \~11,000 attributed to the bottom of Terrace’s level 5, and \~15,600 attributed to the top of level 5 and level 4E. Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes. Retouched tools are dominated by splintered pieces, endscrapers and retouched blanks, with Vale Comprido points (the Proto-Solutrean fossil director) being barely inexistent in chert and mostly produced in dolerite (Belmiro et al., 2021).

Previous studies showed that the main raw materials used at the site are chert, quartz, greywacke and smaller amounts of chalcedony, with the percentages of these changing across occupations. Notably, the lower Proto-Solutrean levels show high frequencies of quartz (Belmiro et al., 2021), whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter (Cascalheira, 2010). Likewise, different raw materials appear to have been used for specific purposes,with chert being particularly associated  with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils (Horta et al., 2019)

Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius. Our previous studies showed, as expected, the existence of local cherts in the assemblages (\@fig-cherts a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site (Belmiro et al., 2025). Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert (Luedtke, 1978). A significant  portion of non-local cherts was also identified (\@fig-cherts b-c), with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain. This indicates varying degrees of contact or long-range mobility throughout the UP, likely influenced by shifts in settlement patterns and the duration of site occupation.

@@

Previous research suggests that both the Terrace and Shelter functioned as seasonal residential camps, repeatedly occupied for extended periods. This interpretation is based on high-density lithic assemblages, evidence of tool manufacture, heat-cracked rocks associated with grease rendering, diverse faunal remains (marine and terrestrial), and personal ornaments and portable art (Manne et al., 2012). This is true for most of the UP occupations, although with some exceptions.

Cascalheira et al. (2017) applied the Whole Assemblage Behavioral Index (WABI) (Riel-Salvatore, 2010; Riel-Salvatore & Barton, 2004) to reconstruct and compare mobility strategies at Vale Boi. By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean (Cascalheira, 2010; Cascalheira et al., 2013) and Magdalenian (Mendonça, 2009) levels, as well as the Terrace’s Gravettian levels (Marreiros et al., 2015), the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations. Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.

A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009). The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).\

@@

The macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025) and can also be found in an online repository (<https://www.doi.org/10.17605/OSF.IO/DBFT2>). This study focused on the UP chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area. The methodology used in the study, dataset and supporting documents can be found in the article and the online repository.

The Solutrean assemblage technological analysis was published by Cascalheira (2009) as part of a Master’s thesis and following publications (e.g., Cascalheira, 2010; Cascalheira et al., 2013), with the study of levels A to C of the Shelter area. The dataset used for this study was not originally published but can be consulted in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).

The Solutrean analysis in this study is limited to the Shelter area due to the inexistence of technological studies for the better preserved Solutrean levels of the Terrace (e.g., level 4). However, and as seen in Belmiro et al. (2025), the raw material patterns seem to show little differences between the Solutrean in both areas; coupled with the similar radiocarbon dates between the areas (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.

The Proto-Solutrean technological analysis was the result of a Master’s thesis (Belmiro, 2020) and following publication (Belmiro et al., 2021), with the study of levels 4E and 5 of the Terrace area. The dataset and analysis can be consulted in the original online repository (<http://www.doi.org/10.17605/OSF.IO/456EG>).

@@

One challenge of integrating datasets collected by different researchers is variability in methodologies and recorded variables. However, this study minimizes such inconsistencies by ensuring methodological alignment across datasets, as all data collection followed standardized attribute-based lithic analysis protocols. The lithic analysis datasets used in this study adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analysed followed those present in specialized literature, such as Brézillon (1968), and Tixier and Inizan (1980), paired with other Palaeolithic lithic attribute analyses (see e.g., Bicho, 1992; Zilhão, 1997; Cascalheira, 2019; Almeida, 2000; Scerri et al., 2014; Tostevin, 2013).

After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. Some differences in the collection process, especially between the study by Cascalheira (2010) and the following studies, required homogenization. For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al. (2021), retouch was identified as Class variable. In this situation, the Solutrean dataset was modified, so that all artifacts with retouch were identified in the Class variable, instead of the original classification. All changes applied to the original data made for homogenisation between datasets can be consulted in the R Studio scripts in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).

@@

The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>). We used the rrtools package by Marwick et al. (2018) to create a research compendium and write a reproducible journal article. The provided files include the complete set of raw data used in the analysis, along with a custom R project (Wickham, 2015) containing the code required to generate all tables and figures. To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY (additional details can be found in Marwick, 2017).

@@

The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, **@fig-proto-plots**, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in **SOM 1 Tables s9-11** and **s12**, respectively.

@@

Our findings confirm initial expectations regarding mobility patterns, raw material use, and technological strategies in southwestern Iberia during the UP. While core reduction differences between local and non-local cherts were anticipated (\@tbl-expectations, expectations a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi. As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy (Kuhn, 2004)—or as trimmed cores and larger stone volumes, aligning with site provisioning models (Kelly, 1988; Roth & Dibble, 1998; Clark, 1987; Kuhn, 2004).

@@

Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site. However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; **SOM 1 Table S13**), which lack associated cores and knapping debris. This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models (Kuhn, 2004).

@@

Core and blank size comparisons provide further insights. Non-local cherts underwent more intensive reduction than local cherts, as indicated by their higher blank-to-core ratios and the significantly smaller size and weight of non-local chert blanks. These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility. In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials (Kuhn, 2004; Roth & Dibble, 1998; Grove et al., 2023; Surovel, 2009). However, our results show that these differences are not limited to long-term occupations and low mobility. High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves (Cascalheira, 2017; Belmiro et al., 2025). Due to this, we may interpret that non-local raw materials were generally more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.

In contrast to blanks, cores exhibited small, though statistically non-significant, differences in size and weight between local and non-local cherts. These results may be explained by a combination of factors, including the availability of local lithic resources, transport, and knapping strategies focused on the intensive reduction and maintenance of non-local raw materials. Belmiro et al. (2023) highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm. Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted, and normally occur fractured in secondary deposits associated to primary outcrops. As a result, the small size of local chert cores may not necessarily indicate intensive reduction but could instead reflect the initial size of available raw material volumes.

@@

The intensity of reduction and maintenance discussed thus far can also be inferred from the frequency and diversity of retouched tools. In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials (Andrefsky, 1994; Bamforth, 1986; Kelly, 1988; Kuhn, 2004; Perlès, 1992).  Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages. This supports the idea that non-local raw materials were actively procured, and that greater effort was invested in tool production (Andrefsky, 1994; Perlès, 1992; Kelly, 1988) and the maintenance of existing artifacts through retouch (Kuhn, 2004).

While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools (Odell, 1981; Shott, 1986; Siegel, 1984). Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages. This suggests that these materials were not only transported as unspecialised tools but as volumes that supported on-site tool production and diversification.

@@

Previous studies emphasize raw material abundance and quality as key factors driving the procurement and use of non-local lithic resources over local ones, shaping their proportions, use, and maintenance. In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools (Andrefsky, 1994; Perlès, 1992; Kelly, 1988). However, quality may become a less significant factor when non-local raw materials are easily accessible (Akerman et al., 2002; Torrence, 1996; Brown, 1999; Oestmo, 2017). Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources (Roth & Dibble, 1998).

Regarding Vale Boi, quality does not appear to be a defining factor in raw material choice, as typologies typically associated with high-quality lithic resources are also present in other materials. For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition. This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality (Schmidt et al., 2016)—have been identified during the Solutrean in European UP contexts (Bachellerie & Schmidt, 2021; Bordes, 1969; Inizan et al., 1976; Tiffagom, 1998).

Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert (Belmiro et al., 2021). This further supports the idea that raw material choice was not strictly based on quality, as these points were made from a variety of materials, including those less commonly associated with formal tools at the site.

@@

Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts (Bicho et al., 2016). This observation further emphasizes the use of diverse chert types, regardless of their quality.

>>>>

The proportions of local vs non-local raw materials discussed thus far have focused greatly on group mobility, which is linked with residential mobility and occupation duration. Residential mobility (e.g., foraging range size and frequency of residential moves) is a defining characteristic of hunter-gatherer societies and plays a central role in their lifeways [@kelly1995]. It is a key factor driving changes in lithic assemblages [@brown1999; @oestmo2017; @roth1998] as it significantly impacts the cost and use of artefacts and raw materials since they affect the relation between the distance of raw material sources to sites [@kuhn2004]. As such, different raw material and technological patterns have been observed between short-term and long-term occupations.

Short-term occupations typically rely on transported (non-local) toolkits [@torrence1983] and raw materials [@surovell2009], show higher reduction intensity, and an increased tools-to-debitage ratio in the assemblages [@grove2023]. Long-term occupations typically rely on provisioning sites with locally available and suitable raw materials, while non-local materials are gradually used and discarded [@grove2023; @surovell2009]. Non-local raw material cores are highly reduced and extensively exploited [@kuhn2004; @roth1998] and exhibit higher blank-to-core ratios [@roth1998].

Altogether, different studies highlight significant variability in lithic assemblages and raw material proportions, reflecting different facets of hunter-gatherer technological organisation. This variability appears between behaviourally and ethnically similar sites [@bamforth1986], across distinct chrono-cultural periods, and even within a single site among various raw materials. To fully understand past hunter-gatherer behaviours, organisation, and lifeways, it is crucial to examine all stages of lithic production— procurement, distribution, use, maintenance, and discard—through interconnected approaches that combine lithic technology and raw material analyses.

@@

Vale Boi is located on the western coast of the Algarve region, southern Portugal, within a small valley extending \~2 km southward to the Atlantic Ocean (\@fig-map a). The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters (\@fig-map b). Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter (\@fig-map c), which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012]. Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data [@belmiro2025], the Terrace and Shelter areas were selected for the current study.

@@

The Terrace area preserves occupations ranging from the UP to the Early Neolithic. While excavations started in 2000, an additional 8-square-meter trench was opened in 2012 to refine the stratigraphic sequence and assess the presence of earlier UP deposits. Between 2012 and 2016, eight primary litho-stratigraphic units were identified. The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) dated between approximately c. 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c. 26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP [@belmiro2021; @cascalheira2017; @cascalheira2013b].

The Shelter area contains four main lithostratigraphic units, with occupations from the Magdalenian, Solutrean, and Gravettian periods. The Magdalenian and Gravettian levels (Z and D, respectively) were partially eroded, while the Solutrean levels represent the most intensive occupation, with three well-preserved archaeological horizons (layers A to C), dated to approximately c. 24-22 ka cal BP. These occupation levels were found beneath blocks of limestone that likely collapsed from the rock shelter ceiling, probably shortly after the LGM [@cascalheira2010; @cascalheira2017].

The site’s lithic assemblages represent all stages of stone tool production, with raw material use varying by occupation. Technological analyses of these assemblages have provided insights into the adaptive strategies of UP hunter-gatherers at Vale Boi, revealing technological shifts across different cultural phases. Given the chosen sample for this study (see **Section 2.2. Datasets and analysis**), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021]. The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.

The lithic analysis of the Solutrean levels from the Shelter (levels A-C) shows a high number of lithic artifacts, with over 22,000 artifacts analysed (including chips which correspond to \~60% of the assemblage and shatter \~17-19%). The volumes introduced into the site were mostly cortical, given the high frequencies of cortex in blanks and retouched tools. The knapping strategies are dominated by unidirectional and bidirectional prismatic cores, with frequent core maintenance products. Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blanks are the most common retouched tools. Stemmed and winged points, laurel leaves, and shouldered points were also identified [@cascalheira2013a].

The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to \~65-70% of the assemblage and shatter \~20%), \~11,000 attributed to the bottom of Terrace’s level 5, and \~15,600 attributed to the top of level 5 and level 4E. Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes. Retouched tools are dominated by splintered pieces, endscrapers and retouched blanks, with Vale Comprido points (the Proto-Solutrean fossil director) being barely inexistent in chert and mostly produced in dolerite [@belmiro2021].

Previous studies showed that the main raw materials used at the site are chert, quartz, greywacke and smaller amounts of chalcedony, with the percentages of these changing across occupations. Notably, the lower Proto-Solutrean levels show high frequencies of quartz [@belmiro2021], whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter [@cascalheira2010]. Likewise, different raw materials appear to have been used for specific purposes,with chert being particularly associated  with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils [@horta2019].

Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius. Our previous studies showed, as expected, the existence of local cherts in the assemblages (\@fig-cherts a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025]. Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert [@luedtke1978]. A significant  portion of non-local cherts was also identified (\@fig-cherts b-c), with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain. This indicates varying degrees of contact or long-range mobility throughout the UP, likely influenced by shifts in settlement patterns and the duration of site occupation.

@@

Previous research suggests that both the Terrace and Shelter functioned as seasonal residential camps, repeatedly occupied for extended periods. This interpretation is based on high-density lithic assemblages, evidence of tool manufacture, heat-cracked rocks associated with grease rendering, diverse faunal remains (marine and terrestrial), and personal ornaments and portable art [@manne2012]. This is true for most of the UP occupations, although with some exceptions.

Cascalheira et al. (2017) applied the Whole Assemblage Behavioral Index (WABI) [@riel-salvatore2004; @riel-salvatore2010] to reconstruct and compare mobility strategies at Vale Boi. By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean [@cascalheira2010; @cascalheira2013a] and Magdalenian [@mendonça2009] levels, as well as the Terrace’s Gravettian levels [@marreiros2015], the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations. Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.

A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi and following other models relating local/non-local raw materials with mobility [@grove2023; @surovell2009]. The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves [@belmiro2025].\

@@

The macroscopic and petrographic data of cherts was previously collected and published by our team [@belmiro2025] and can also be found in an online repository (<https://www.doi.org/10.17605/OSF.IO/DBFT2>). This study focused on the UP chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area. The methodology used in the study, dataset and supporting documents can be found in the article and the online repository.

The Solutrean assemblage technological analysis was published by Cascalheira [-@cascalheira2009] as part of a Master’s thesis and following publications [e.g., @cascalheira2010; @cascalheira2013a], with the study of levels A to C of the Shelter area. The dataset used for this study was not originally published but can be consulted in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).

The Solutrean analysis in this study is limited to the Shelter area due to the inexistence of technological studies for the better preserved Solutrean levels of the Terrace (e.g., level 4). However, and as seen in [@belmiro2025], the raw material patterns seem to show little differences between the Solutrean in both areas; coupled with the similar radiocarbon dates between the areas (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi.

The Proto-Solutrean technological analysis was the result of a Master’s thesis [@belmiro2020] and following publication [@belmiro2021], with the study of levels 4E and 5 of the Terrace area. The dataset and analysis can be consulted in the original online repository (<http://www.doi.org/10.17605/OSF.IO/456EG>).

@@

One challenge of integrating datasets collected by different researchers is variability in methodologies and recorded variables. However, this study minimizes such inconsistencies by ensuring methodological alignment across datasets, as all data collection followed standardized attribute-based lithic analysis protocols. The lithic analysis datasets used in this study adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analysed followed those present in specialized literature, such as Brézillon [-@brézillon1968], and [@tixier1980], paired with other Palaeolithic lithic attribute analyses [e.g., @bicho1992; @zilhão1997; @cascalheira2019; @tostevin2012].

After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. Some differences in the collection process, especially between the study by Cascalheira [-@cascalheira2010] and the following studies, required homogenization. For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al. [-@belmiro2021], retouch was identified as Class variable. In this situation, the Solutrean dataset was modified, so that all artifacts with retouch were identified in the Class variable, instead of the original classification. All changes applied to the original data made for homogenisation between datasets can be consulted in the R Studio scripts in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).

@@

The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>). We used the rrtools package by Marwick et al. [-@marwick2018] to create a research compendium and write a reproducible journal article. The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickham2015] containing the code required to generate all tables and figures. To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY (additional details can be found in @Marwick2017).

@@

The sizes of blanks and cores were explored through scatterplots of width and length (\@fig-gravettian-plots**, @fig-proto-plots**, \@fig-solutrean-plots a) and boxplots for weight (\@fig-gravettian-plots, @fig-proto-plots, \@fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in **SOM 1 Tables s9-11** and **s12**, respectively.

@@

Our findings confirm initial expectations regarding mobility patterns, raw material use, and technological strategies in southwestern Iberia during the UP. While core reduction differences between local and non-local cherts were anticipated (\@tbl-expectations, expectations a-b), these differences were largely shaped by transportation strategies used to bring non-local cherts to Vale Boi. As such, the occurrence of non-local chert cores at Vale Boi depended on whether these materials arrived as finished tools or blanks—likely reflecting an individual provisioning strategy [@kuhn2004]—or as trimmed cores and larger stone volumes, aligning with site provisioning models [@kelly1988; @roth1998; @clark1987; @kuhn2004] .

@@

Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site. However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; **SOM 1 Table S13**), which lack associated cores and knapping debris. This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models [@kuhn2004].

@@

Core and blank size comparisons provide further insights. Non-local cherts underwent more intensive reduction than local cherts, as indicated by their higher blank-to-core ratios and the significantly smaller size and weight of non-local chert blanks. These results align with previous studies, which propose that the intensive reduction of cores—manifested in smaller core sizes, smaller debitage products, and a higher amount of debitage per core—may be linked to lower mobility. In this context, lower mobility results in reduced access to non-local raw materials, which are therefore more intensively reduced and eventually replaced by local raw materials [@kuhn2004; @roth1998; @grove2023; @surovell2009]. However, our results show that these differences are not limited to long-term occupations and low mobility. High blank-to-core ratios and smaller blank sizes in non-local cherts are also observed in the Gravettian occupation, previously interpreted as a short-term occupation with frequent moves [@cascalheira2017a; @belmiro2025]. Due to this, we may interpret that non-local raw materials were generally more intensively reduced than local cherts throughout all the UP, independent of the duration of the occupation.

In contrast to blanks, cores exhibited small, though statistically non-significant, differences in size and weight between local and non-local cherts. These results may be explained by a combination of factors, including the availability of local lithic resources, transport, and knapping strategies focused on the intensive reduction and maintenance of non-local raw materials. Belmiro et al. [-@belmiro2023] highlight that chert nodules in the Algarve region vary between \~5-10 cm, with a maximum diameter of 20 cm. Although abundant and easily accessible in exposed cliffs, local chert nodules are generally small and prone to fracturing when extracted, and normally occur fractured in secondary deposits associated to primary outcrops. As a result, the small size of local chert cores may not necessarily indicate intensive reduction but could instead reflect the initial size of available raw material volumes.

@@

The intensity of reduction and maintenance discussed thus far can also be inferred from the frequency and diversity of retouched tools. In fact, retouched tools are often key elements used in the literature to distinguish between differential use of local and non-local raw materials [@andrefsky1994; @bamforth1986; @kelly1988; @kuhn2004; @perlès1992]. Our results align with expected outcomes, as non-local cherts exhibit higher frequencies of retouched tools and higher tool-to-blank ratios across all assemblages. This supports the idea that non-local raw materials were actively procured, and that greater effort was invested in tool production [@andrefsky1994; @perlès1992; @kelly1988] and the maintenance of existing artifacts through retouch [@kuhn2004].

While the high intensity of retouch in non-local cherts aligns with expectations (\@tbl-expectations, expectations 1c and 2c), their unexpectedly high tool diversity contradicts common mobility models which expect mobility to be linked with limited tool variability and an increase of easily reconfigurable tools [@odell1981; @shott1986; @siegel1984]. Non-local cherts at Vale Boi contain a diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages. This suggests that these materials were not only transported as unspecialised tools but as volumes that supported on-site tool production and diversification.

@@

Previous studies emphasize raw material abundance and quality as key factors driving the procurement and use of non-local lithic resources over local ones, shaping their proportions, use, and maintenance. In this context, the scarcity or low quality of local resources would drive the procurement of non-local materials and require more effort in preparing transportable cores and producing tools [@andrefsky1994; @perlès1992; @kelly1988]. However, quality may become a less significant factor when non-local raw materials are easily accessible [@akerman2002; @torrence1986; @brown1999; @oestmo2017]. Even in the presence of high-quality local materials, effort was still invested in transporting non-local resources [@roth1998].

Regarding Vale Boi, quality does not appear to be a defining factor in raw material choice, as typologies typically associated with high-quality lithic resources are also present in other materials. For example, Solutrean bifacial tools and points were produced using Lower Jurassic and Triassic local chert, which may be considered lower quality than Rio Maior Cretaceous chert due to its coarser texture and less homogeneous composition. This suggests that the choice of raw material was not solely dependent on quality, especially when techniques such as heat treatment—the process of heating lithic raw material to improve knapping quality [@schmidt2016]—have been identified during the Solutrean in European UP contexts [@bachellerie2022; @bordes1969; @inizan1976; @tiffagom1998].

Similarly, Vale Comprido points, a fossil indicator of the Proto-Solutrean, were primarily made from other lithic resources (e.g., dolerite), with only two identified points made of chalcedony and chert [@belmiro2021]. This further supports the idea that raw material choice was not strictly based on quality, as these points were made from a variety of materials, including those less commonly associated with formal tools at the site.

@@

Finally, a visual inspection of a cache found in the Shelter area attributed to the Gravettian revealed a high frequency of backed bladelets made from both local and non-local cherts [@bicho2016]. This observation further emphasizes the use of diverse chert types, regardless of their quality.
